---
title: "Installing arenalytics offline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing arenalytics offline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This guide explains how to install arenalytics on a machine that has **no
internet connection**. You will need the file
**`arenalytics_offline_pkgs.zip`** provided by your administrator or a
colleague (see the section *Preparing the bundle* at the bottom if you are the
one creating it).

---

## Prerequisites

Make sure the following software is already installed on your machine before
you start.

| Software | Notes |
|---|---|
| **R ≥ 4.1** | <https://cran.r-project.org> |
| **RStudio** | <https://posit.co/download/rstudio-desktop/> |
| **Rtools** *(Windows only)* | Required to compile packages from source. Download the version matching your R installation from <https://cran.r-project.org/bin/windows/Rtools/> |
| **Xcode Command Line Tools** *(macOS only)* | Open the Terminal app and run `xcode-select --install` |
| **Quarto CLI** | Required for report generation. <https://quarto.org/docs/get-started/> |

---

## Step 1 — Create a new R project

1. Open **RStudio**.
2. Go to **File › New Project › New Directory › New Project**.
3. Name the project (e.g. `arenalytics`) and choose a folder on your machine.
4. Click **Create Project**.

You should now be inside a fresh RStudio project with an empty workspace.

---

## Step 2 — Copy and extract the bundle

1. Copy **`arenalytics_offline_pkgs.zip`** into the project folder you just
   created (the same folder that contains the `.Rproj` file).
2. Extract it there:
   - **Windows**: right-click the zip → *Extract All…* → keep the default
     destination.
   - **macOS / Linux**: double-click the zip, or run in the Terminal:

```bash
unzip arenalytics_offline_pkgs.zip -d arenalytics_offline_pkgs
```

You should now have a sub-folder called **`arenalytics_offline_pkgs/`**
containing many `.tar.gz` files and a `PACKAGES` index file.

---

## Step 3 — Install the dependencies

Open the RStudio **Console** pane and run the following commands.

### 3a. Point R to the local package repository

```{r local-repo}
local_repo <- paste0("file:///", normalizePath("arenalytics_offline_pkgs", winslash = "/"))
```

### 3b. Install all required packages

```{r install-deps}
install.packages(
  pkgs = c(
    "bsicons", "bslib", "DT", "dplyr", "ggplot2", "htmltools",
    "jsonlite", "purrr", "quarto", "rlang", "scales", "shiny",
    "shiny.i18n", "shinyjs", "shinyWidgets", "srvyr",
    "stringr", "tibble", "tidyr", "zip"
  ),
  repos        = local_repo,
  type         = "source",
  dependencies = TRUE
)
```

> This step compiles packages from source and may take **5–15 minutes**.
> You will see a lot of output in the console — that is normal.

---

## Step 4 — Install arenalytics

Copy the **`arenalytics`** package source folder (the folder that contains the
`DESCRIPTION` file) into your project directory, then run:

```{r install-pkg}
install.packages("arenalytics", repos = NULL, type = "source")
```

If you received a pre-built `.tar.gz` of the package instead, pass its path
directly:

```{r install-tarball}
install.packages("path/to/arenalytics_1.0.0.tar.gz", repos = NULL, type = "source")
```

---

## Step 5 — Launch

```{r launch}
library(arenalytics)
shiny_run_arenalytics()
```

The arenalytics dashboard should open in your browser.

---

## Troubleshooting

**`package 'XYZ' is not available`**  
The package was not included in the bundle. Ask your administrator to
regenerate the bundle using `tools/make_offline_bundle.R`.

**Compilation errors on Windows**  
Rtools is likely missing or the wrong version. Make sure the Rtools version
matches your R version (e.g. Rtools44 for R 4.4.x).

**Compilation errors on macOS**  
Run `xcode-select --install` in the Terminal and retry.

**`could not find function "shiny_run_arenalytics"`**  
The arenalytics package itself was not installed. Repeat Step 4.

---

## Preparing the bundle *(for administrators)*

If you are the one providing the zip file, run the following script on a
machine **with** internet access. It downloads all required packages and their
dependencies into a single zip ready to share:

```r
source("tools/make_offline_bundle.R")
```

This creates `arenalytics_offline_pkgs.zip` in your working directory.
Share that zip along with this vignette (or `tools/install_offline.md`) with
your users.
